Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2018-08-01 paperIEEE.tex
Reviewer two comments
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@140 db7c848b-7755-4df2-b763-8796797afa52

The Equalizer parallel rendering framework as first presented in~\cite{EMP:09} has
demonstrated its general versatility and the usefulness of its minimally-intrusive
programming design in a variety of applications and projects.
In particular, the integration of large-scale parallel rendering algorithms,
APIs for developing complex distributed applications, and many
individual features make Equalizer a unique, open source framework to develop
visualization applications for virtually any type of setup and use case.
While individual applications and case studies using the framework, as well as
new algorithms and system components extending it have been presented
since the initial release of Equalizer, many important features and functionalities
have not previously been published and are presented here.
In this report, we thus present an updated comprehensive review of the
integration of research, application use cases and commercial developments
with respect to Equalizer as well as novel comparative experimental results
of its scalability features.

@@

The remainder of this paper is structured as follows: First we provide an update
on related work since the introduction of \textsf{Equalizer}. The main body of this paper
then presents new performance features, VR algorithms, usability
features to build complex applications, main novel features of the underlying
\textsf{Collage} network library and a quick overview of the main
\textsf{Equalizer}-based applications. A result section presents new experiments
not previously published, followed by the discussion and conclusion.

@@

In 2009 we presented \textsf{Equalizer}~\cite{EMP:09}, which introduced the
architecture of a generic parallel rendering framework and summarized our work
in parallel rendering. Since then, an extensive Programming and User Guide
provides in-depth documentation on using and programming
\textsf{Equalizer}~\cite{Eilemann:13}. In the following related work we assume
these two publications and their references as a starting point, and focus on
the new work published since 2009.

@@

@@

% This section introduces new decomposition modes, runtime adaptations and other
% optimizations to increase the rendering performance for a wide variety of
% algorithms and use cases.

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first (2D),
sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the
following we present the newly added decomposition modes and motivate their use
case, which bring the overall feature set way beyond the typical sort-first and
sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
modes. The compound concept to set up scalable rendering is presented in
\cite{EMP:09}.

@@

Sort-first (\fig{fig:loadeq}) and sort-last load balancing are the most obvious
optimizations for these parallel rendering modes. Our load equalizers are fully
transparent for application developers; that is, they use a reactive approach
based on past rendering times. This assumes a reasonable frame-to-frame
coherence. Equalizer implements two different algorithms, a
\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. The result section
provides some evidence on the strengths and weaknesses of both algorithms.

@@

Cross-segment load balancing (\fig{fig:vieweq}) addresses the optimal resource
allocation of $n$ rendering resources to $m$ output channels (with $n\geq m$). A
view equalizer works in conjunction with load equalizers balancing the
individual output channels. It monitors the usage of shared source channels
across outputs and activates them to balance the rendering time of all outputs.
In \cite{EEP:11}, we provide a detailed description and evaluation of our
algorithm.

@@

The region of interest is the screen-space 2D bounding box enclosing the data
rendered by a single resource. We have extended the core parallel rendering
framework to use an application-provided ROI to optimize the
\textsf{load\_equalizer} as well as the image compositing. The load equalizer
uses the ROI to refine its load grid to the regions containing data. The
compositing code uses the ROI to minimize image readback and network
transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
the algorithm, and show that using ROI can quadruple the rendering performance,
in particular for the costly compositing step in sort-last rendering.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
most applications and large data scenarios. To easily customize the
synchronization of all threads on a single process, \textsf{Equalizer}
implements three threading modes:  Full synchronization, draw synchronization
and asynchronous. Note that the execution between node processes is always
asynchronous, for up to \textsf{latency} frames.

@@

%------------------------------------------------------------------------------
\section{Virtual Reality Features}
%------------------------------------------------------------------------------

Virtual Reality is an important field for parallel rendering. It does however
require special attention to support it as a first-class citizen in a generic
parallel rendering framework. \textsf{Equalizer} has been used in many virtual
reality installations, such as the Cave2 \cite{FNTTL:13}, the high-resolution C6
CAVE at the KAUST visualization laboratory, and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations, motivated by application use cases.

@@

Head tracking is the minimal feature needed to support immersive installations.
\textsf{Equalizer} does support multiple, independent tracked views through the
observer abstraction (\sref{sec:observer}). Built-in VRPN support enables the
direct, application-transparent configuration of a VRPN tracker device.
Alternatively, applications can provide a $4\times 4$ tracking matrix. Both
CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
displays are implemented.

@@

%------------------------------------------------------------------------------
\subsection{Physical and Logical Visualization Setup}

@@

%----------------------------------------------------------------------
\section{Example Applications} \label{sApplications}
%----------------------------------------------------------------------

@@

RTT Deltagen (now Dassault 3D Excite) is a commercial application for
interactive, high quality rendering of CAD data. The RTT Scale module,
delivering multi-GPU and distributed execution, is based on \textsf{Equalizer}
and \textsf{Collage}, and has driven many of the aforementioned features.

@@

RTNeuron \cite{HBBES:13} is a scalable real-time rendering tool for the
visualisation of neuronal simulations based on cable models. It is based on
OpenSceneGraph for data management and Equalizer for parallel rendering, and
focuses not only on fast rendering times, but also on fast loading times with no
offline preprocessing. It provides level of detail (LOD) rendering, high quality
anti-aliasing based on jittered frusta and accumulation during still views,
interactive modification of the visual representation of neurons on a per-neuron
basis (full neuron vs. soma only, branch pruning depending on the branch level,
\dots). RTNeuron implements both sort-first and sort-last rendering with order
independent transparency.

>>>>

The Equalizer parallel rendering framework, first presented in~\cite{EMP:09},
has demonstrated its general versatility and the usefulness of its
minimally-intrusive design in a variety of applications and projects. In
particular, the integration of large-scale parallel rendering algorithms, APIs
for developing complex distributed applications, and many individual features
make Equalizer a unique, open source framework to develop visualization
applications for virtually any type of setup and use case. While individual
applications and case studies using the framework, as well as new algorithms
and system components extending it have been presented since the initial
release of Equalizer, many features and functionalities have not been published
yet and are presented here. In this report, we thus present an updated
comprehensive review of the integration of research, application use cases and
commercial developments with respect to Equalizer as well as novel comparative
experimental results of its scalability features.

@@

The remainder of this paper is structured as follows: First we provide an
update on related work since the introduction of \textsf{Equalizer}. The main
body of this paper then presents new performance features, VR algorithms,
usability features to build complex applications, an overview of the underlying
\textsf{Collage} network library and a summary of the main
\textsf{Equalizer}-based applications. A result section presents new
experiments, followed by the discussion and conclusion.

@@

In 2009 we presented \textsf{Equalizer}~\cite{EMP:09}, which introduced the
architecture of a generic parallel rendering framework and summarized our work
in parallel rendering. Since then, an extensive Programming and User Guide
provides in-depth documentation on using and programming
\textsf{Equalizer}~\cite{Eilemann:13}. We assume these two publications and
their references as a starting point, and focus on the new work published since
2009.

@@

\section{Fundamentals}

The core task of developing an Equalizer application is to separate the
rendering code from the application logic. The application main thread will
execute its main loop as before, but instead of issuing rendering calls it will
request a frame to be rendered from Equalizer. Depending on the configuration,
Equalizer will then decompose and synchronize the rendering tasks. The
application rendering code is executed in parallel under a given render
context, which is the core entity abstracting the application-specific
rendering algorithm from the system-specific configuration. This render context
formalizes the minimal abstraction for parallel rendering, and specifies:

\begin{compactdesc}

 \item[Buffer] OpenGL-style read and draw buffer as well as color mask. These
 parameters are influenced by the current eye pass, eye separation and
 anaglyphic stereo settings.

 \item[Viewport] Two-dimensional pixel viewport restricting the rendering area.
 The pixel viewport is influenced by the destination viewport and viewports set
 for sort-first decompositions.

 \item[Frustum] Frustum parameters as defined by \textsf{glFrustum}. Typically
 the frustum is used to set up the OpenGL projection matrix. The frustum is
 influenced by the destinations view definition, sort-first, pixel and subpixel
 decomposition, tracking head matrix and the current eye pass.

 \item[Head Transformation] A transformation matrix positioning the frustum.
 For planar views this is an identity matrix. In immersive rendering, it is
 normally used to set up the `view' part of the modelview matrix, before static
 light sources are defined.

 \item[Range] A one-dimensional range with the interval [0..1]. This parameter is
 optional and should be used by the application to render only the appropriate
 subset of its data for sort-last rendering.

\item[View] The view object from the logical rendering rendering configuration
(\sref{sView}). Holds view-specific data, such as camera, model or any other
application state.

\end{compactdesc}

An Equalizer configuration determines how the application is executed in
parallel. It has three parts: A resource section describing the physical
rendering resources (node, pipe, window, channel), a section describing the
visualization setup (\sref{sView}), and a compound section describing how these
resources are to be used for rendering. Compounds describe the rendering task
decomposition and result recomposition, and their algorithm has been introduced
in \cite{EMP:09}.

Our execution model is fully asynchronuous, and only introduces synchronization
points when strictly needed. The main synchronization points are: configured
swap barriers between a set of output channels which have to display
simultaneously, the availability of input frames for scalable rendering, and a
task synchronization to prevent runaway of the main loop execution. By default,
Equalizer keeps up to one frame of latency in execution, that is, some resource
might render the next frame while others are still finishing the current.
Nonetheless, resources which are ready will immediately display their result.
This asynchronous execution architecture, coupled with a frame of latency,
allows pipelining of many operations, such as the application event processing,
task computation and load balancing, rendering, image readback, compression,
network transmission, and compositing. It also hides small imbalances in the
task distribution, as they can average out over multiple frames. It is
fundamental to provide optimal performance, in particular for larger cluster
sizes.

@@

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first
(2D), sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}.
In the following we present new decomposition modes and motivate their use
case, which bring the overall feature set way beyond the typical sort-first and
sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
modes. The compound concept to set up scalable rendering is presented in
\cite{EMP:09}.

@@

Sort-first (\fig{fig:loadeq}) and sort-last load balancing are the most obvious
optimizations for these parallel rendering modes. Our load equalizers are fully
transparent for application developers; that is, they use a reactive approach
based on past rendering times. This assumes a reasonable frame-to-frame
coherence. Equalizer implements two different algorithms, a
\textsf{load\_equalizer} and a \textsf{tree\_equalizer}. Both implementations
modify the \textsf{viewport} or \textsf{range} of their direct child compounds.
The result section provides some evidence on the strengths and weaknesses of
both algorithms.

@@

Cross-segment load balancing (CSLB, \fig{fig:vieweq}) optimizes resource
allocation of $n$ rendering resources to $m$ output channels (with $n\geq m$).
A view equalizer works in conjunction with load equalizers balancing the
individual output channels. It monitors the usage of shared source channels
across outputs and activates them to balance the rendering time of all outputs.
CSLB is unique in that it supports load balancing for planar and non-planar
visualization systems, as well as a flexible allocation of rendering resources
to individual output channels. In \cite{EEP:11}, we provide a detailed
description and evaluation of our algorithm and show that it significantly
improve performance over a static allocation.

@@

The region of interest is the screen-space 2D bounding box enclosing the data
rendered by a single resource. We use an application-provided ROI to optimize
the \textsf{load\_equalizer} as well as image compositing in our framework. The
load equalizer uses the ROI to refine its load grid to the regions containing
data. The compositing code uses the ROI to minimize image readback and network
transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
the algorithm, and show that using ROI can quadruple the rendering performance,
in particular for the costly compositing step in sort-last rendering.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
most applications and large data scenarios. We implement three threading modes:
Full synchronization, draw synchronization and asynchronous. Note that the
execution between node processes is always asynchronous, for up to
\textsf{latency} frames.

@@

%------------------------------------------------------------------------------
\section{Virtual Reality}
%------------------------------------------------------------------------------

Virtual Reality is an important field for parallel rendering, which requires
special attention to support it as a first-class citizen in a generic parallel
rendering framework. \textsf{Equalizer} has been used in many virtual reality
installations, such as the Cave2 \cite{FNTTL:13}, the high-resolution C6 CAVE
at the KAUST visualization laboratory, and head-mounted displays
(\fig{FIG_teaser}). In the following we lay out the features needed to support
these installations.

@@

Head tracking is fundamental to support immersive installations. We support
multiple, independent tracked views through the observer abstraction
(\sref{sec:observer}). Built-in VRPN support enables direct,
application-transparent configuration of a VRPN tracker device. Alternatively,
applications can provide a $4\times 4$ tracking matrix. Both CAVE-like tracking
with fixed projection surfaces and HMD tracking with moving displays are
implemented.

@@

%------------------------------------------------------------------------------
\subsection{Physical and Logical Visualization Setup}\label{sView}

@@

%----------------------------------------------------------------------
\section{Applications} \label{sApplications}
%----------------------------------------------------------------------

@@

RTT Deltagen (\fig{fLayout}, now Dassault 3D Excite) is a commercial
application for interactive, high quality rendering of CAD data. The RTT Scale
module, delivering multi-GPU and distributed execution, is based on
\textsf{Equalizer} and \textsf{Collage}, and has driven many of the
aforementioned features.

@@

RTNeuron \cite{HBBES:13} is a scalable real-time rendering tool for the
visualisation of neuronal simulations based on cable models (\fig{FIG_teaser}
(a)). It is based on OpenSceneGraph for data management and Equalizer for
parallel rendering, and focuses not only on fast rendering times, but also on
fast loading times with no offline preprocessing. It provides level of detail
(LOD) rendering, high quality anti-aliasing based on jittered frusta and
accumulation during still views, interactive modification of the visual
representation of neurons on a per-neuron basis (full neuron vs. soma only,
branch pruning depending on the branch level, \dots). RTNeuron implements both
sort-first and sort-last rendering with order independent transparency.
